\section{Introduction}

Garbage collection refers to the process of automatically managing
heap allocated memory on behalf of the running process by
identifying parts of memory that are no longer needed. \cite[Introduction]{go_gcguide_2022}
This is often performed by the runtime of a programming language while the
program is executing. \cite[Introduction]{go_spec_2023}

Most programming languages allocate values with static
lifetimes\footnote{variable available for the whole runtime of the program
\cite[Abstract]{static-dynamic-scope_tanter_2009}} in main memory along with
the executable code. Values that are alive for a certain scope are allocated
using the call stack\footnote{stores information about running subroutines /
functions\cite[2.2 Call Stacks]{call-stack_mcmaster-memon_2006}} without
requiring dynamic allocation. These Variables can't escape the scope they were
defined in and must be dynamically allocated if accessing them outside of
their scope is desired.

This requires the programmer to allocate and deallocate these
variables to prevent memory leaks\footnote{allocated no longer needed memory
not deallocated \cite[1.2.1 A Practical Object Ownership
Model]{practical_heine-lam_2003}} provided the programming language does not
perform garbage collection. 

\begin{listing}[H] 
    \begin{minted}{c} 
typedef struct { 
    char *name;
    double age; 
} Person;

Person *new_person(char *name, double age) {
  Person *p = malloc(sizeof(Person));
  p->age = age, p->name = name;
  return p;
}
    \end{minted}
    \caption{C heap allocation example}
    \label{code:c_heap_alloc_example}
\end{listing}

Listing \autoref{code:c_heap_alloc_example} showcases a possible use case for
dynamic memory allocation. The \texttt{Person} structure is filled with values
defined in the parameters of the \texttt{new\_person} function. This structure,
if stack allocated, would not live longer than the scope of the
\texttt{new\_person} function, thus rendering this function useless. To create
and use a \texttt{Person} structure outside of its scope, the structure has to
be dynamically allocated via the \texttt{malloc} function defined in the
\mintinline{c}{#include <stdlib.h>} header.

% TODO: showcase free and issues occurring:
% - use after free
% - memory leak
% cite OWASP
% Show example with go, escape to heap showcase
