\chapter{Comparison with other Memory Management Techniques}

\section{Manual Memory Management}

\section{Lifetimes and Borrow Checking}

The desire for the performance of manual memory management and the safety of garbage collection has led to the development
of a new memory management technique called \textit{lifetimes and borrow checking}.
The main idea behind this technique is that the corresponding \texttt{free} calls for heap memory can be automatically inserted
at compile time by the compiler, if the compiler can prove that the memory is no longer needed.
When a variable is no longer needed, it is said to have reached the end of its \textit{lifetime} hence the name of the technique.
% TODO: introduce borrow checking here shortly for context

This memory management technique in the presented form was first introduced in the Rust programming language
\cite[1. Introduction]{rust_borrow_formalism_2021}.
Because Rust was the first language to implement this concept, the examples in this section will be written in Rust.

\subsection{Ownership}

The first step to understand this technique is to understand the concept of \textit{ownership}.

In Rust, every value is always owned by exactly a variable inside a scope.
When the variable goes out of scope, the value is dropped.
The ownership of a value can be transferred to another variable by \textit{moving} it.
This can be either in the form of an assignment or as a function return value.
When a value is moved, the previous owner can no longer access the value.
When a value is dropped by going out of scope, any memory it owns is freed, including heap memory \cite[59-61]{rust_book_2019}.

A major contrast in Rust compared to other programming languages like C
is that variable assignments like \texttt{let a = b} are moving the value
instead of copying it \cite[2.2 Ownership]{rust_borrow_formalism_2021}. The same goes for the parameter values for function calls.
Because of this variables can not be used after being used in a variable assignment or function call.

A example showcasing the ownership concept similar
\footnote{
    The presented Rust example differs from the C example because the
    \texttt{create\_person} function does not return a reference but a value.
    However the struct consists of a \texttt{String} which is a heap allocated
    dynamic length string that gets allocated by the \texttt{.into()} call
    converting the static \texttt{\&str} into \texttt{String}, so 
    the example still requires the used heap memory to be freed.
    References/Borrows to temporary values are not allowed in Rust so
    the only way to force a heap allocation of the whole struct would
    be to use a \texttt{std::boxed::Box<T>}. This was not used in this
    example for legibility reasons.
}
to the C example presented in the introduction
can be found in \autoref{code:rust_ownership_person_example}
\footnote{
    A more idomatic Rust implementation would define the new and print functions as methods of the Person struct.
    This was not done here to keep the example simple for readers not familiar with Rust.
}.

\begin{listing}[H] 
    \begin{minted}{rust}
struct Person {
    name: String,
    age: f64,
}

fn new_person(name: String, age: f64) -> Person {
    Person { name, age }
}

fn print_person(person: Person) {
    println!("{} is {} years old.", person.name, person.age);
}

fn main() {
    let person = new_person("Rainer Zufall".into(), 42.0);
    let person1 = person; // value of person is moved to person1
    // print_person(person); // error: use of moved value: `person`
    print_person(person1);

    { // Example of sub-scope
        let person2 = new_person("Jona Zufall".into(), 13.0);
        print_person(person2);
    } // person2 is dropped here
} // person1 and person are dropped here
    \end{minted}
    \caption{Person struct example in Rust demonstrating ownership}
    \label{code:rust_ownership_person_example}
\end{listing}

\subsection{Borrowing}

The second step to understand this technique is to understand the concept of \textit{borrowing}.
Allowing only one owner to access a variable at a time would be too restrictive
for many uses cases.

As a example calling the \texttt{print\_person} function twice on the same person
would not be possible, because the ownership of the person would be moved to the
function after the first call and cannot be accessed anymore.
This is illustrated in \autoref{code:rust_person_double_print}.

\begin{listing}[H] 
    \begin{minted}{rust}
let person = new_person("Rainer Zufall".into(), 42.0);
print_person(person);
print_person(person); // error: use of moved value: `person`
    \end{minted}
    \caption{Failed attempt to print a person twice in Rust due to lost ownership}
    \label{code:rust_person_double_print}
\end{listing}

The solution to this problem is the concept of borrowing.
It essentially is the pointer concept from C and other languages but with
the ownership model of Rust in mind, which imposes some restrictions on it.
Borrowing allows a value owner to give another function or struct
access to a value without giving ownership to the function \cite[2.3 Borrowing]{rust_borrow_formalism_2021}.
A owner can hand out many read-only borrows to a value at the same time,
but only one mutable borrow at a time. This is done to avoid data races but
is not strictly needed for the memory management aspect of the technique \cite[90]{rust_book_2019}.

A modified version of the presented person example to make use of borrows
can be found in \autoref{code:rust_borrow_person_example}.

\begin{listing}[H] 
    \begin{minted}{rust}
fn print_person(person: &Person) {
    println!("{} is {} years old.", person.name, person.age);
}

fn main() {
    let person = new_person("Rainer Zufall".into(), 42.0);
    print_person(&person); // borrow using &
    print_person(&person); // borrow a second time
} // person is dropped here
    \end{minted}
    \caption{Person struct example in Rust demonstrating borrowing}
    \label{code:rust_borrow_person_example}
\end{listing}

% TODO: explain borrow checking
% TODO: lifetimes for functions returning values and structs

\subsection{Multi-owner values using reference counters}

Some usecases require a value to be owned by multiple owners at the same time.
These usecases include shared memory and cyclic data structures.

Allowing multiple owners for a single value can be done by using reference counters as a escape hatch.
A reference counter is a data structure that keeps track of the number of owners
of a value and drops the value when the number of owners reaches zero.
This is done by incrementing the counter when a new owner is created using \texttt{.clone()}
and decrementing it when a owner is dropped.
Reference counter implementations are available in the Rust standard library
as \texttt{std::rc::Rc<T>} \cite[320-323]{rust_book_2019} and \texttt{std::sync::Arc<T>} \cite[361]{rust_book_2019} for single threaded and atomic
multi-threaded use respectively.
An example of this can be found in \autoref{code:rust_ref_counter_example}.

\begin{listing}
    \begin{minted}{rust}
use std::rc::Rc;

fn main() {
    // Create an Rc that contains a person
    let person = Rc::new(new_person("Rainer Zufall".into(), 42.0));

    // Clone the Rc to create additional references
    // These can be moved to other owners and outlive the original Rc instance
    let clone1 = Rc::clone(&person);
    let clone2 = Rc::clone(&person);

    println!("Reference count of person: {}", Rc::strong_count(&person));

    // Access the data through the cloned references
    println!("clone1 data: {:?}", clone1);
    println!("clone2 data: {:?}", clone2);

    // When the references go out of scope, the reference count decreases
    drop(clone1);
    println!("Count after dropping clone1: {}", Rc::strong_count(&person));

    drop(clone2);
    println!("Count after dropping clone2: {}", Rc::strong_count(&person));

    // At this point, the reference count drops to zero, and the memory is
    // deallocated because the last reference is dropped.
}

// output:
// Reference count of person: 3
// clone1 data: Person { name: "Rainer Zufall", age: 42.0 }
// clone2 data: Person { name: "Rainer Zufall", age: 42.0 }
// Count after dropping clone1: 2
// Count after dropping clone2: 1
    \end{minted}
    \caption{Reference counter example in Rust}
    \label{code:rust_ref_counter_example}
\end{listing}

The cloned reference counter instance can be moved to other owners and outlive the original instance.
This works great for non-cyclic data structures, but not so well by itself for cyclic data structures
(see \autoref{sec:reference_counting_cycles}) because the reference counter will never reach zero
and the memory will never be freed.
To solve this problem, you can use weak references \cite[334-335]{rust_book_2019} or break the cycle manually
when you are done with the data structure.
When the developer does not deal with this problem a memory leak will occur.


